Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-08-16 reviewers_response.md

> 2c. Opaque code is presented in order to generate plots, e.g.
```{r}
palDF <- ceObj@clusterLegend[[1]]
pal <- palDF[, "color"]
names(pal) <- palDF[, "name"]
pal["-1"] = "transparent"
plot(fit$points, col = pal[primaryClusterNamed(ceObj)], main = "", pch = 20, xlab = "Component1", ylab = "Component2") legend(x = "topleft", legend = names(pal), cex = .5, fill = pal, title = "Sample")
```

> While this complexity may be necessary, perhaps some of it could be encapsulated as accessor functions in the package? Too much complexity here may cause users to miss the forest for the trees.

We have created the `plotReduceDims()` function in `clusterExperiment` to address this issue.

> The authors could better motivate (or at least explain the impact of) some of the default parameters and procedures.
> Why do we set a zcut threshold of 3 for the `scone` filtering? 
> Why K=50 for zinbwave?
> RSEC parameters
>
> How should a user decide on a value for these parameters?

### 2018-08-16 workflow/workflow.Rmd
Initial draft for review

  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages. 
keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
editor_options: 
  chunk_output_type: console
---

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and the identification of differentially expressed (DE) genes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true, with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for dowstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.

One of the cornerstones of Bioconductor is the use of common data structures, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
Throughout the workflow, we use a single object of the `SingleCellExperiment` class to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. 
The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data (see Figure 2 of [@huber2015orchestrating] for an overview of `SummarizedExperiment`).
The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].

@@

The scRNA-seq dataset we use as a case study was generated to study the differentitation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Risso2018; @Street2017].

It was found that the first bifurcation produces two lineage trajectories: the first one (purple curve in Figure \@ref(fig:stemcelldiff)) generates mature sustentacular cells (mSUS) through a differentiation process characterized by the absence of cell cycle and division. The other lineage trajectory (blue and orange curves in Figure \@ref(fig:stemcelldiff)) generates proliferative GBCs, which are characterized by cell cycle and division. These, in turn, generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve).

In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes x cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

Note that the preprocessing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper. These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g, 3' end sequencing, droplet-based methods versus full-length mRNA, plate-based methods). 
One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files. More details can be found in the `scPipe` vignette, available at https://bioconductor.org/packages/scPipe.

@@

The source code to reproduce the analysis in the workflow are available at https://github.com/fperraudeau/singlecellworkflow. The packages used in the workflow can be installed using the following in R:

```{r, eval=FALSE}
install.packages(c("doParallel", "gam", "RColorBrewer"))

install.packages("BiocManager")

BiocManager::install(c("scone", "zinbwave", "clusterExperiment", "slingshot",
                       "BiocParallel"))
```

@@

```{r packages}
# Bioconductor
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)
library(slingshot)

# CRAN
library(doParallel)
library(gam)
library(RColorBrewer)

@@

urls = c("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE95601&format=file&file=GSE95601%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz",
         "https://raw.githubusercontent.com/rufletch/p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt")

@@

As discussed abouve, we store the data in an object of the class `SingleCellExperiment`. In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object. Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.

@@

In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER). This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that this is not an endogenous gene.

@@

@@

Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects". Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch. Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing]. In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data is processed, or different times at which samples were collected.

@@

```{r spikebatch, fig.cap="Distribution of reads mapped onto ERCC spike-ins per batch."}
boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
        outline = 0, ylab="Expression measure", las = 3)
```

Figure \@ref(fig:spikebatch) shows how for some batches no reads are mapped to ERCC spike-ins, presumably because the spike-ins were not included. Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.

@@

In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 151 cells not assigned to any cluster (i.e., cluster label of `-2`).

@@

One drawback of this strategy is that the logarithm is not a variance-stabilizing transformation. Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package. And these may be a better choice in real data analyses. Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function. Here, we show an example with the `vst` function from `DESeq2`.

```{r, eval=FALSE}
library(DESeq2)
dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = vst,
                           filterNames = "var_vst")
dataObj <- filterData(dataObj, percentile=1000, filterStats="var_vst")
```

@@

<!-- An example of how to integrate that strategy in this workflow is provided here. Note that to follow `scran` recommendations we first normalize the data using the `scran` method. -->

<!-- ```{r, eval=FALSE} -->
<!-- library(scran) -->
<!-- dataScran <- computeSumFactors(dataObj) -->
<!-- dataScran <- normalize(dataScran) -->

<!-- tv <- trendVar(dataScran, use.spikes = FALSE) -->
<!-- dv <- decomposeVar(dataScran, fit = tv) -->
<!-- ``` -->

<!-- The results of `dv$bio` can then be used to sort the genes and keep the top 1,000. -->

DR: I'm not sure whether to keep this last example using scran. It's more complicated because we cannot directly use the clusterExperiment functions and it's somewhat involved (we would need to explain what the size factors are etc.). Also, the results make no sense (negative variances!). Perhaps we should just mention the method without the code?

@@

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor R package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a cell-level intercept, which serves as a global-scaling normalization factor. The user can also specify both gene-level and cell-level covariates. The inclusion of observed and unobserved cell-level covariates enables normalization for complex, non-linear effects (often referred to as batch effects), while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.png")
```

@@

As with most dimensionality reduction methods, the user needs to specify the number of factors (`K`, the number of columns of `W`), which correspond to the dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`. Note that if the users include more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.

```{r zinb,eval=FALSE}
print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
                                      residuals = TRUE, 
                                      normalizedValues = TRUE)))
```

```{r}
#save(dataObj, file= 'dataObj_afterZinbwave_k10.rda')
load('dataObj_afterZinbwave.rda')
```

@@

The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that, in this case, the low-dimensional matrix `W` is not included in the computation of residuals to avoid the removal of the biological signal of interest.

@@

We are going to set these normalized values to be our first assay, so that they become the default in future plots.

```{r setNormAssay}
assays(dataObj)<-assays(dataObj)[c("normalizedValues","residuals","counts","weights")]
```

@@

@@

Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, which is stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object. This matrix can be retrieved using the `reducedDim()` extractor.

@@

In this workflow, the user-supplied dimension `K` of the low-dimensional space is set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

@@

Here, we fit a few different models, specifying a different value of `K`, using the `zinbFit` function, which, unlike `zinbwave`, returns the estiamtes of all the parameters of the model, as well as the value of the likelihood function.

```{r, eval=FALSE}
k_values <- c(1: 5, 10)

fit <- lapply(k_values, function(k) zinbFit(dataObj, K = k, which_assay = "counts"))
saveRDS(fit, file = "Fit_allK.rds")
```

```{r}
k_values <- c(1: 5, 10)
fit <- readRDS("fit_allK.rds")
plot(k_values, sapply(fit, zinbBIC, t(counts(dataObj))), 
     ylab = "BIC", xlab = "K", type = "b")
plot(k_values, sapply(fit, zinbAIC, t(counts(dataObj))),
     ylab = "AIC", xlab = "K", type = "b")
```

In this case, the BIC suggests a value of `K = 3`, while the AIC suggests a value of `K >= 10`.

Note, that since ZINB-WaVE is linear in the number of genes and number of cells, and quadratic on `K`, fitting multiple models and comparing their goodness-of-fit is infeasible for large datasets. In such cases, choosing a single value of `K` may be needed. In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets.

DR: There is quite a difference between AIC and BIC in this case, and it is actually hard to argue one value over another... how should we discussed these results? Also, if I keep going AIC will keep decreasing...

@@

```{r rsec_50,eval=FALSE}
print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, reduceMethod="zinbwave",
                                clusterFunction = "hierarchical01",
                                ncores = NCORES,
                                dendroReduce="zinbwave",
                                subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                          clusterArgs=list(nstart=10)),
                                consensusProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                consensusMinSize = 10)))
```

```{r}
#save(dataObj, file = 'dataObj_afterRSEC_k10.rda')
load('dataObj_afterRSEC.rda')
```

As one can see from the code above, there are many parameters that a user can set in RSEC. Many of the values that we choose are the default values of the function, but we nonetheless specify them for clarity and reproducibility. Here, we briefly describethe role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].

- `k0s`: the values of `k` to start the sequential algorithm.
- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster.
- `mergeMethod`: our consensus clustering may find too many small clusters. If that's the case a merging step can be employed to merge clusters that do not show enough differential expression. Here, we skip this step by setting the method as "none".
- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).

@@

The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examinemakeConsensus)), where columns correspond to cells and rows to different clusterings. Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap accross rows. The first row correspond to a consensus clustering across all candidate clusterings.

@@

Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is anologous to Figure \@ref(fig:mdsW) for the original published clusters.

@@

This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object. The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (ie. a matrix and a vector) or, as below, components of a `SingleCellExperiment` object. When a `SingleCellExperiment` object is provided as input, the ouput will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function. For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.

@@

@@

pseudoCe1 <- pseudoCe[,!is.na(t)]
orderSamples(pseudoCe1)<-order(t[!is.na(t)])

>>>>

  - name: Fanny Perraudeau
    affiliation: 
      - Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
      - Whole Biome, Inc., San Francisco, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages. 
keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
editor_options: 
  chunk_output_type: console
---

@@

Single-cell RNA sequencing (scRNA-seq) is a powerful and promising class of high-throughput assays that enable researchers to measure genome-wide transcription levels at the resolution of single cells. To properly account for features specific to scRNA-seq, such as zero inflation and high levels of technical noise, several novel statistical methods have been developed to tackle questions that include normalization, dimensionality reduction, clustering, the inference of cell lineages and pseudotimes, and the identification of differentially expressed (DE) genes. While each individual method is useful on its own for addressing a specific question, there is an increasing need for workflows that integrate these tools to yield a seamless scRNA-seq data analysis pipeline. This is all the more true with novel sequencing technologies that allow an increasing number of cells to be sequenced in each run. For example, the Chromium Single Cell 3' Solution was recently used to sequence and profile about 1.3 million cells from embryonic mouse brains.

@@

Here, we propose an integrated workflow for downstream analysis, with the following four main steps: (1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates, using the `zinbwave` Bioconductor package; (2) robust and stable cell clustering using resampling-based sequential ensemble clustering, as implemented in the `clusterExperiment` Bioconductor package; (3) inference of cell lineages and ordering of the cells by developmental progression along lineages, using the `slingshot` Bioconductor package; and (4) DE analysis along lineages.

One of the cornerstones of Bioconductor is the use of common data containers, implemented as classes in the S4 object-oriented system of R [@huber2015orchestrating].
Throughout the workflow, we use a single object of the `SingleCellExperiment` class to store the scRNA-seq data along with any gene or cell-level metadata available from the experiment. 
The `SingleCellExperiment` class [@SingleCellExperiment] is an extension of the `SummarizedExperiment` class [@SummarizedExperiment] specifically designed for single-cell data (see Figure 2 of [@huber2015orchestrating] for an overview of `SummarizedExperiment`).
The use of the `SingleCellExperiment` class allows us to seamlessly transition between the steps of this workflow, as well as to integrate this workflow with the upstream QC and EDA steps provided in [@Lun2016].

@@

The scRNA-seq dataset we use as a case study was generated to study the differentiation of HBC stem cells into different cell types present in the olfactory epithelium. To map the developmental trajectories of the multiple cell lineages arising from HBCs, scRNA-seq was performed on FACS-purified cells using the Fluidigm C1 microfluidics cell capture platform followed by Illumina sequencing. The expression level of each gene in a given cell was quantified by counting the total number of reads mapping to it. Cells were then assigned to different lineages using a statistical analysis pipeline analogous to that in the present workflow. Finally, results were validated experimentally using in vivo lineage tracing. Details on data generation and statistical methods are available in [@Fletcher2017; @Risso2017; @Risso2018; @Street2017].

It was found that the first bifurcation produces two lineage trajectories: the first one (purple curve in Figure \@ref(fig:stemcelldiff)) generates mature sustentacular cells (mSUS) through a differentiation process characterized by the absence of cell cycle and division. The other lineage trajectory (blue and orange curves in Figure \@ref(fig:stemcelldiff)) generates proliferative GBCs, which are characterized by cell cycle and division. This, in turn, diverges into two lineages that generate mature olfactory sensory neurons (mOSN; orange curve) and microvillous cells (MV; blue curve), respectively.

In this workflow, we describe a sequence of steps to recover the lineages found in the original study, starting from the genes by cells matrix of raw counts publicly-available at https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE95601.

Note that the preprocessing steps needed to produce such a count matrix from the raw FASTQ files are outside of the scope of this paper. These are very important steps that should be taken with care and the optimal workflow depends on the technology and the protocol used to collect the data (e.g, 3' end sequencing, droplet-based methods versus full-length mRNA, plate-based methods). 
One way to carry out these steps within the Bioconductor framework is the `scPipe` package [@tian2017scpipe], which provides functions to quantify gene expression starting from raw sequencing files. More details can be found in the `scPipe` vignette and in [@tian2017scpipe].

@@

The source code to reproduce the analysis in the workflow is available at https://github.com/fperraudeau/singlecellworkflow. The packages used in the workflow can be installed using the following in R:

```{r, eval=FALSE}
# CRAN
install.packages(c("doParallel", "gam", "RColorBrewer"))

# Bioconductor
install.packages("BiocManager")
BiocManager::install(c("scone", "zinbwave", "clusterExperiment", "slingshot",
                       "BiocParallel", "DESeq2", "scater"))
```

@@

```{r packages}
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)
library(slingshot)
library(doParallel)
library(gam)
library(RColorBrewer)

@@

urls = c(paste0("https://www.ncbi.nlm.nih.gov/geo/download/",
                "?acc=GSE95601&format=file&file=GSE95601",
                "%5FoeHBCdiff%5FCufflinks%5FeSet%2ERda%2Egz"),
         paste0("https://raw.githubusercontent.com/rufletch/",
                "p63-HBC-diff/master/ref/oeHBCdiff_clusterLabels.txt"))

@@

As discussed above, we store the data in an object of the class `SingleCellExperiment`. In addition to keeping the expression values and their associated metadata within a single object, `SingleCellExperiment` allows the user to store normalization factors and the results of dimensionality reduction methods (such as `zinbwave`) within the object. Throughout the workflow, we will highlight when this special features of `SingleCellExperiment` are used.

@@

In this case, we specify two different sets of spike-ins, the ERCC spike-ins and the transgenic genes (eGFP and Cre-ER). This is a slight abuse of notation, as the CreER and eGFP genes are not exactly spike-ins, but it is nonetheless useful to specify that these are not endogenous genes.

@@

Note that here we use the function `counts()` from the `SingleCellExperiment` package that retrieves the matrix stored in the assay slot named `counts`. This is equivalent to `assay(dataObj, "counts")` but is obviously more convenient when accessing counts multiple times. It is hence a good idea to name the assay slot that stores the original counts "counts", as done above.

@@

Cells were processed in 18 different runs of the Fluidigm C1 system: we refer to each run as a "batch", and we refer to the technical differences between each run as "batch effects". Note that in this particular case, the batch structure is essentially defined by the limitation of the C1 system to process up to 96 cells at a time and it is hence natural to identify each C1 run as one batch. Batch effects are ubiquitous in high-throughput genomics [@leek2010tackling] and single-cell RNA-seq is not an exception [@hicks2017missing]. In other contexts, depending on experimental design, batches may correspond to different library preparations, different labs in which the data are processed, or different times at which samples are collected.

@@

```{r spikebatch, fig.cap="Distribution of reads mapped onto ERCC spike-ins per batch."}
boxplot(colData(dataObj)$ERCC_reads~batch, col = col_batch,
        outline = 0, ylab="Reads mapped to ERCC", las = 3)
```

Figure \@ref(fig:spikebatch) shows how for some batches no reads were mapped to ERCC spike-ins, presumably because the spike-ins were not included. Since the utility of the spike-ins is limited in this example, we remove the spike-ins from our object.

@@

In the original work [@Fletcher2017], cells were clustered into 14 different clusters, with 233 cells not assigned to any cluster (i.e., cluster label of `-2`).

@@

One drawback of this strategy is that the logarithm is not a variance-stabilizing transformation. Variance stabilizing transformations based on the negative binomial distribution are available, e.g., in the `DESeq2` package, and may be a better choice in real data analyses. Note that the `makeFilterStats` function of `clusterExperiment` can be used with any user-supplied function. Here, we show an example with the `vst` function from `DESeq2`.

```{r, eval=FALSE}
## Not run
library(DESeq2)
dataObj <- makeFilterStats(dataObj, filterStats = "var", transFun = vst,
                           filterNames = "var_vst")
dataObj <- filterData(dataObj, percentile=1000, filterStats="var_vst")
```

@@

@@

Here, we perform dimensionality reduction using the zero-inflated negative binomial-based wanted variation extraction (ZINB-WaVE) method implemented in the Bioconductor package `zinbwave`. The method fits a ZINB model that accounts for zero inflation (dropouts), over-dispersion, and the count nature of the data. The model can include a cell-level intercept, which serves as a global-scaling normalization factor. The user can also specify both gene-level and cell-level covariates. The inclusion of cell-level covariates enables normalization for batch effects, while gene-level covariates may be used to adjust for sequence composition effects (e.g., gene length and GC-content effects). A schematic view of the ZINB-WaVE model is provided in Figure \@ref(fig:zinbschema). For greater detail about the ZINB-WaVE model and estimation procedure, please refer to the original manuscript [@Risso2017].

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2018)."}
knitr::include_graphics("zinb_schema.png")
```

@@

As with most dimensionality reduction methods, the user needs to specify the number of factors (`K`, the number of columns of `W`), which correspond to the dimensions for the new low-dimensional space. Here, we use `K = 50` dimensions and adjust for batch effects via the matrix `X`. Note that if the user includes more genes in the analysis, it may be preferable to reduce `K` to achieve a similar computational time.

```{r zinb}
print(system.time(dataObj <- zinbwave(dataObj, K = 50, X = "~ Batch",
                                      residuals = TRUE,
                                      normalizedValues = TRUE)))
```

@@

The function `zinbwave` returns a `SingleCellExperiment` object that includes the normalized expression measures (`notmalizedValues`), defined as deviance residuals from the fit of the ZINB-WaVE model with user-specified gene- and cell-level covariates. Such residuals can be used for visualization purposes (e.g., in heatmaps, boxplots). Note that the low-dimensional matrix `W` is not included in the computation of the residuals to avoid the removal of the biological signal of interest.

@@

@@

```{r}
assayNames(dataObj)
```

We set the normalized values as our first assay, so that they become the default in the plots below.

```{r setNormAssay}
assays(dataObj) <- assays(dataObj)[c("normalizedValues", "counts", "weights",
                                     "residuals")]
```

@@

Perhaps the most important component of the results of `zinbwave` is the low-dimensional matrix `W`, stored as the `zinbwave` component of the `reducedDims` slot of the `SingleCellExperiment` object. This matrix contain the biological signal of interest and can be retrieved using the `reducedDim()` extractor.

@@

In this workflow, the user-supplied dimension `K` of the low-dimensional space was set to `K = 50`. The resulting low-dimensional matrix `W` can be visualized in two dimensions by performing multi-dimensional scaling (MDS) using the Euclidian distance. To verify that `W` indeed captures the biological signal of interest, we display the MDS results in a scatterplot with colors corresponding to the original published clusters (Figure \@ref(fig:mdsW)).

@@

Here, we fit a few different models, specifying a different value of `K`, using the `zinbFit` function, which, unlike `zinbwave`, returns the estimates of all the parameters of the model, as well as the value of the likelihood function.

```{r}
k_values <- c(1:5, 10)

fit <- lapply(k_values, function(k) zinbFit(dataObj, K = k, which_assay = "counts"))
```

```{r bic, fig.cap="BIC of the model with different values of K."}
k_values <- c(1: 5, 10)
plot(k_values, sapply(fit, zinbBIC, t(counts(dataObj))), 
     ylab = "BIC", xlab = "K", type = "b")
```

```{r aic, fig.cap="AIC of the model with different values of K."}
plot(k_values, sapply(fit, zinbAIC, t(counts(dataObj))),
     ylab = "AIC", xlab = "K", type = "b")
```

In this case, the BIC suggests a value of `K = 3` (Figure \@ref(fig:bic))), while the AIC suggests a value of `K >= 10` (Figure \@ref(fig:aic))).

Note, that since ZINB-WaVE is linear in the number of genes and number of cells, and quadratic on `K`, fitting multiple models and comparing their goodness-of-fit may not be feasible for large datasets. In such cases, choosing a single value of `K` may be needed. In practice, we found that `K` between 10 and 50 are reasonable choices in real datasets.

@@

```{r rsec_50}
print(system.time(dataObj <- RSEC(dataObj, k0s = 4:15, alphas = c(0.1),
                                betas = 0.8, reduceMethod="zinbwave",
                                clusterFunction = "hierarchical01",
                                ncores = NCORES,
                                dendroReduce="zinbwave",
                                subsampleArgs = list(resamp.num=100,
                                           clusterFunction="kmeans",
                                          clusterArgs=list(nstart=10)),
                                consensusProportion = 0.7,
                                mergeMethod = "none", random.seed=424242,
                                consensusMinSize = 10)))
```

As one can see from the code above, there are many parameters that a user can set in RSEC. Many of the values that we choose are the default values of the function, but we nonetheless specify them for clarity and reproducibility. Here, we briefly describe the role of the parameters that we set, but a more comprehensive explanation is provided in the function's man page (`?RSEC`) and in [@Risso2018].

- `k0s`: the values of `k` to start the sequential algorithm.
- `alphas`: similarity required in the clustering of the subsampling similarity matrix (larger alphas are less stringent).
- `betas`: similarity required to deem a cluster "stable" in sequential clustering (larger betas require greater similarity).
- `clusterFunction`: function used for the subsample clustering: `hierarchical01` means that we use hierarchical clustering on a dissimilarity matrix based on the co-clustering proportions across subsamples.
- `dendroReduce`: the dimensionality reduction to use for the creation of the dendrogram of the found clusters.
- `subsampleArgs`: a list of parameters for the subsample clustering; here we specified the base clustering function as k-means, the number of resamplings and the number of starting values for the k-means algorithm.
- `consensusProportion`: the proportion of times that two sets of samples should be together in order to be grouped into a cluster in the consensus clustering step.
- `mergeMethod`: our consensus clustering may find too many small clusters. If that's the case a merging step can be employed to merge clusters that do not show enough differential expression. Here, we skip this step by setting the method as "none".
- `consensusMinSize`: minimum size required for a set of samples to be considered in a cluster; clusters smaller than this number will be removed and the corresponding samples will be set to "unclustered" (`-1`).

The results of `RSEC` is an object of the `ClusterExperiment` class. This class extends the `SingleCellExperiment` class, by adding additional slots to store the clustering results. In addition, it has an informative `show` method that displays important information on the clustering.

Note that an object of class `ClusterExperiment` is also of class `SingleCellExperiment`. Hence, all the methods define for `SingleCellExperiment` will still work on it.

```{r}
is(dataObj, "SingleCellExperiment")
```

@@

The resulting candidate clusterings can be visualized using the `plotClusters` function (Figure \@ref(fig:examinemakeConsensus)), where columns correspond to cells and rows to different clusterings. Each sample is color-coded based on its clustering for that row, where the colors have been chosen to try to match up clusters that show large overlap across rows. The first row correspond to a consensus clustering across all candidate clusterings.

@@

Finally, we can visualize the cells in a two-dimensional space using the MDS of the low-dimensional matrix `W` and coloring the cells according to their newly-found RSEC clusters (Figure \@ref(fig:mdsWce)); this is analogous to Figure \@ref(fig:mdsW) for the original published clusters.

@@

This analysis is performed by the `slingshot` function and the results are stored in a `SlingshotDataSet` object. The minimal input to this function is a low-dimensional representation of the cells and a set of cluster labels; these can be separate objects (i.e. a matrix and a vector) or, as below, components of a `SingleCellExperiment` object. When a `SingleCellExperiment` object is provided as input, the output will be an updated `SingleCellExperiment` object containing a `SlingshotDataSet` as an element of the `int_metadata` list, which can be accessed through the `SlingshotDataSet` function. For more low-level control of the lineage inference procedure, the two steps of fitting the MST and the simultaneous principal curves may be run separately via the functions `getLineages` and `getCurves`.

@@

The `show` method of the `SlingshotDataSet` class displays useful information on the inferred lineages.

@@

dataFilt1 <- dataFilt[,!is.na(t)]
orderSamples(dataFilt1)<-order(t[!is.na(t)])
